---
title: "Introducing the BGFanalyzeR Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing the BGFanalyzeR Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The package adds the concept of microbial fermentation data to the R programming language. It was developed for **B**io**G**as **F**ermentations, but concepts and workflows might also be applied to other types fermentations.

The central concept of this package is a novel S3 class object, the `BGF` which allows to store and visualize experimental data, and also supports experiment design.

This vignette at first provides some background on what biogas fermentations are, as this might be a rather exotic topic to most R users.
Then, the `BGF`-concept will be announced and some easy methods how to set up a `BGF` are shown.

However, this vignette focuses on the basics and other vignettes provide more detailed explanations for distinct `BGF`-types and experimental workflow.

```{r package schema,echo = FALSE,fig.height=7,fig.width=7}
library(ggplot2)
# Setup and scale ####
dimensi=c(-25:60)
xlimit=c(-10,40)
ylimit=dimensi
basefont=4
fontstep=1
pointSize=40

Plot<-
# Generate an empty scaled plot ####
ggplot(mapping = aes(dimensi,dimensi))+
  scale_x_continuous(limits=xlimit)+
  #scale_y_continuous(limits=ylimit)+
  theme_void()+

  # R environment frame ####
  geom_rect(aes(xmin=8.5,xmax=37.5,ymin=7,ymax=33),color="steelblue",alpha=.0009,linewidth=1.5)+
  geom_text(mapping = aes(x=23,y=30,label="R programming language"),size=basefont,color="steelblue")+

  ## Box 1 BGFanalyzeR ####
  geom_rect(aes(xmin=10,xmax=20,ymin=8,ymax=27),fill="orange",alpha=.02,color="black")+
  geom_text(mapping = aes(x=15,y=25,label="BGFanalyzeR"),size=basefont,color="white")+

  geom_rect(aes(xmin=11,xmax=19,ymin=11,ymax=16),fill="orange",color="black")+
  geom_text(mapping = aes(x=15,y=13.7,label="data analysis/\nprocessing"),size=(basefont-fontstep),color="white")+

  geom_rect(aes(xmin=11,xmax=19,ymin=17,ymax=22),fill="orange",color="black")+
  geom_text(mapping = aes(x=15,y=19.7,label="experiment\ndesign"),size=(basefont-fontstep),color="white")+

  ## Box 2 other packages ####
  geom_rect(aes(xmin=26,xmax=36,ymin=8,ymax=27),fill="orange",alpha=.02,color="black")+
  geom_text(mapping = aes(x=31,y=25,label="other packages"),size=basefont,color="white")+

  geom_rect(aes(xmin=27,xmax=35,ymin=9,ymax=13),fill="orange",color="black")+
  geom_text(mapping = aes(x=31,y=11,label="keras3"),size=basefont,color="white")+


  geom_rect(aes(xmin=27,xmax=35,ymin=14,ymax=18),fill="orange",color="black")+
  geom_text(mapping = aes(x=31,y=16,label="phyloseq"),size=basefont,color="white")+

  geom_rect(aes(xmin=27,xmax=35,ymin=19,ymax=23),fill="orange",color="black")+
  geom_text(mapping = aes(x=31,y=21,label="Biogas"),size=basefont,color="white")+

  ## BGF_test ####
  geom_rect(aes(xmin=19.5,xmax=24.5,ymin=14,ymax=20),fill="black")+
  geom_polygon(mapping = aes(x=c(24,24,26.5),y=c(11.5,22.5,17)),fill="black")+
  geom_text(mapping = aes(x=22.7,y=17.1,label="BGF"),size=basefont,color="white")+

  # R Output
  geom_point(aes(x=17,y=-14),size=pointSize,color="darkgreen",alpha=.4)+
  geom_text(mapping = aes(x=17,y=-14,label="standardized data\nformats (.csv/ .RDS)"),size=(basefont-fontstep),color="white")+

  geom_point(aes(x=2,y=-14),size=pointSize,color="darkgreen",alpha=.4)+
  geom_text(mapping = aes(x=2,y=-14,label="Plots"),size=(basefont-fontstep),color="white")+

  geom_point(aes(x=32,y=-14),size=pointSize,color="darkgreen",alpha=.4)+
  geom_text(mapping = aes(x=32,y=-14,label="further analysis"),size=(basefont-fontstep),color="white")+

  # idea ####
  geom_rect(aes(xmin=-7,xmax=0,ymin=2,ymax=38),fill="darkblue",alpha=.02)+
  geom_text(mapping = aes(x=-3.5,y=19,label="Idea /\nResearch question"),size=basefont,color="white",angle=90)+

  # experiments ####
  geom_rect(aes(xmin=0,xmax=39,ymin=46,ymax=53),fill="darkgreen")+
  geom_text(mapping = aes(x=20,y=50,label="Lab experiments /\nSeries of fermentations"),size=basefont,color="white")+

  # connections ####
  ## Idea to BGFanalyzeR ####
  geom_line(mapping = aes(x=c(-.5,10.5),y=c(19,19)),linewidth=2.3,arrow=arrow(angle = 30, length = unit(0.25, "inches"),
                                                                              ends = "both", type = "open"))+
  geom_text(aes(x=5,y=20.5,label="conception"),size=basefont)+

  ## Idea to Lab experiment ####
  geom_line(mapping = aes(x=c(-3.5,0.5),y=c(37.5,49.5)),linewidth=2.3,arrow=arrow(angle = 30, length = unit(0.25, "inches"),

                                                                                                                                ends = "last", type = "open"))+
  geom_text(aes(x=-3.,y=43.,label="leads to"),angle=55,size=basefont)+

  ## BGFanalyzeR to Plots ####
  geom_line(mapping = aes(x=c(17,2),y=c(8.5,-9)),linewidth=2.3,arrow=arrow(angle = 30, length = unit(0.25, "inches"),
                                                                                ends = "first", type = "open"))+
  geom_text(aes(x=8.5,y=1.5,label="visualize"),angle=33,size=basefont)+

  ## BGFanalyzeR to data formats ####
  geom_line(mapping = aes(x=c(17,17),y=c(8.5,-10)),linewidth=2.3,arrow=arrow(angle = 30, length = unit(0.25, "inches"),
                                                                                ends = "last", type = "open"))+
  geom_text(aes(x=17,y=-1.,label="data storage /\nexchange"),angle=90,size=basefont)+

  ## other packages to further analysis ####
  geom_line(mapping = aes(x=c(32,32),y=c(8.5,-9)),linewidth=2.3,arrow=arrow(angle = 30, length = unit(0.25, "inches"),
                                                                         ends = "last", type = "open"))+
  geom_text(aes(x=31,y=1.5,label="allows"),angle=90,size=basefont)+

  ## Lab experiments to BGFanalyzeR ####
  geom_line(mapping = aes(x=c(12,12),y=c(46.5,26.5)),linewidth=2.3,arrow=arrow(angle = 30, length = unit(0.25, "inches"),
                                                                          ends = "last", type = "open"))+
  geom_text(aes(x=12,y=36.5,label="Input\ndata"),angle=90,size=basefont)


Plot
```

## Background - A Microbial Biogas Fermentation
In the sense of the `BGFanalyzer` package, a biogas fermentation is the anaerobic breakdown of organic material by microorganisms into gaseous end products.

The most industrialized version of a biogas fermentation might be bio methane production, also known as anaerobic digestion.
During this process, several microorganisms from the bacterial and archaeal domain of life break down sugars, proteins and fatty acids to terminally form methane (CH~4~) out of it.

Despite methane, biological fermentations can also yield in the formation of other (valuable) gases, such as hydrogen (H~2~) or CO~2~. 
Although only CH~4~ is commonly referred to as 'biogas', all microbial mediated processes that lead to a target gas formation are considered as 'biogas' fermentation in the sense of the `BGFanalyzer` package.

## The `BGF` Concept
The heart of the `BGFanalyzer` package is the new `BGF` object class, which allows to store experimental data of several fermentations.
The package is independent of the 'experimental hardware' and can be used to build `BGF`'s from external text files or interactively in R.

### `BGF` structure
A `BGF` is an S3 data object with a `list` like structure.
Each `BGF` has three standard layers.
These are:

* `ExpParam`: a `list` to store information on the `BGF` itselfe, like its name (`ExpParam$name`) or its measurement type (`ExpParam$MeasurementType`), or information all fermentaions of the `BGF` (should) share like the process temperature (`ExpParam$ProcessTemp`) or the inoculum to substrate ratio (`ExpParam$InocToSubRatio`). 

* `metaData`: a `data.frame` that contains information specific for each fermentation in a `BGF`. 
Three columns exist by default. Each row represents a single fermentation. 
The first column holds the so called reactor layout (`metaData$Layout`), a user specified grouping variable that needs to be specified for each fermentation in a `BGF`. 
The second column can be used to mark a fermentation as blank run (`metaData$Blank`). If a 'Blank' exists, it is possible to subtract its biogas production from other fermentations. 
The last column is used to exclude a fermentation from the set (`metaData$Excluded`).
Excluded fermentations can be filtered quickly from plots and layout specific yield calculations. 
Further columns can be added by the user using dedicated functions. 
For each fermentation in the `BGF`, one row in the `metaData`-layer must exist. 

* `BioGasData`: a `data.frame` used to store the experimental data of each fermentation. 
It has seven default columns and during object creation one empty row per specified fermentation is added. 
Each row is a fermentation-specific observation. 
The first column holds a specific ID (`BioGasData$reactor`), that links a fermentation to the `metaData`-layer of a `BGF`. 
The second column stores the fermentation time (`BioGasData$time`), which is the time that has passed since a fermentation has started. 
This means it is a time difference and can have any appropriate unit (e.g. "hours","days","weeks",etc.). 
The third column is meant to store the amount of fermentation product (`BioGasData$product`), usually an accumulating exhaust gas volume in terms of a biogas fermentation. 
The fourth column referrers to the product generation (`BioGasData$production`), e.g the amount of new product that has formed since the last observation. 
The fifth column is used to calculate the amount of net product(`BioGasData$netProduct`), a specifically corrected amount of the `$product` column. 
The sixth column stores the yield of the fermentation (`BioGasData$yield`), which means how many product was formed per input. 
Finally, the seventh column is used to calculate the relative production, e.g. the ratio product/product~final~ in percent. 
It is also possible to add new columns to that layer, if additional parameters, e.g. the pH value, for the fermentations are recorded.

The decision to make the `BGF` an S3 type object was based on two considerations.

First, S3 is the most primitive object-orientated programming system in R and most of the objects and functions R beginners encounter are of type S3.
We wanted to make working with a `BGF`'s as easy as possible for everyone and thus the S3 system seemed like the right choice as it's familiar to R beginners and experts as well.

Second, the S3 system offers a high degree of flexibility, so every user is free to add new layers to the `BGF` as needed. 
One could imagine to add a fourth layer with e.g. sequencing results of microbial community of individual fermentations.
However, the focus of the `BGFanalyzeR` package is the analysis biogas fermentation data and the general advice is to not change the three-layer architecture of a `BGF`.

## Building a `BGF`

The simplest way for an R user to create a `BGF` is the `BGF()` function. It requires only the `ReactorLayout` argument to be set and will then create a BGF.

```{r first-BGF}
library(BGFanalyzeR)

# create an example BGF with five fermentations having the reactor layout "A", "B", "C", "D", "E"
myBGF <- BGF(ReactorLayout = LETTERS[1:5])

# inspect the BGF
myBGF

```
When a `BGF` is printed to the R console, a textual summary is shown.
In the first line, the name of the `BGF` and the number of fermentations within is displayed.
For the above example, the name is 'new_BGF' and consists of 5 fermentations.

The name of a `BGF` can be set during its creation with `BGF()` and other helper functions:

```{r a named-BGF}
# create a named BGF by specifying the 'name' argument in BGF()
namedBGF <- BGF(ReactorLayout = paste0("2*",LETTERS[1:5]),
                name = "named BGF")

# inspect namedBGF
namedBGF
```

The name of the new `BGF` is 'named BGF' and it contains 10 fermentations, instead.
Note, that a shortcut for setting multiple identical reactor layouts was used during object creation of 'namedBGF' (type `?correct_RLayout` in the R console for details).

Alternatively, a new name can be set using the function `alter_whatever()` or `base` R syntax:

```{r re-naming-BGF}
# rename myBGF using alter_whatever()
myBGF <- alter_whatever(myBGF,
                        layer = "ExpParam",
                        what = "name",
                        value = "new name")

# rename namedBGF using base syntax
namedBGF$ExpParam$name <- "new name"

# do myBGF and namedBGF have the same name now?
myBGF$ExpParam$name == namedBGF$ExpParam$name

# inspect renaming results
get_whatever(myBGF,
             layer = "ExpParam",
             what = "name")

# or
namedBGF$ExpParam$name
```

Following the first line with name and number of fermentations of a `BGF`, comes a block providing the number of variables in each layer of a `BGF`.
In case of 'myBGF' and 'namedBGF', both have only the default variables set in each layer.
This means, the `ExpParam`-layer has four 'experimental parameters' (`ExpParam$name`,`ExpParam$InocToSubRatio`,`ExpParam$ProcessTemp`,`ExpParam$MeasurementType`), the `metaData`-layer has three 'meta variables' (`metaData$Layout`, `metaData$Blank`, `metaData$Excluded`) and the `BioGasData`-layer has seven 'fermentation variables' (`BioGasData$reactor`, `BioGasData$time`, `BioGasData$product`, `BioGasData$production`, `BioGasData$net_product`, `BioGasData$yield`, `BioGasData$rel_production`). 

At this point the number of rows in the `metaData`- and `BioGasData`-layer are identical, but all columns in the `BioGasData`-layer contain only `NA` values.

Let's take a closer look whats already inside our `BGF` objects:

```{r inspect-ExpParam}

# list 'ExpParam'-layers of both BGF's
# myBGF
myBGF$ExpParam

# namedBGF
namedBGF$ExpParam

```

The `ExpParam`-layers of both `BGF`'s are identical, as we have not specified `BioGasData$InocToSubRatio`, `BioGasData$ProcessTemp` or `BioGasData$MeasurementType` manually during object creation, and renamed both objects to 'new name' afterwards.
Thus all parameters despite the name are on their default value.

Now we do the same for the `metaData`-layer:

```{r inspect-metaData}
# list 'metaData'-layers 
# myBGF
myBGF$metaData

# namedBGF
namedBGF$metaData

```

As we can see, both have identical columns (`Layout`, `Blank`, `Excluded`) but differ in the number of rows.
In addition, the `metaData`-layer has row names, by default 'Rn' where 'n' is an increasing integer number.

Whether this 'R' is for 'row', 'replicate', 'run', 'reactor' or 'relation' remains unknown.
However, these row names must match the unique identifiers used in `BioGasData$reactor` of a `BGF`. 

While row names serve as unique identifiers for a fermentation in the `metaData`-layer, its `Layout` column is used to group fermentations according a user specified variable. 
As we have initially specified five, and ten reactor layouts during the creation of 'myBGF' and 'namedBGF', respectively, these columns are not identical for the two `BGF`'s.

For both `BGF`'s, the `Blank` and `Excluded` columns contain only a `logic` `FALSE`, as none of the fermentations in either `BGF` was classified as 'Blank' or 'Excluded' so far.

Finally, let's compare the `BioGasData`-layers:

```{r inspect-BioGasData}
# list 'BioGasData'-layers
# myBGF
myBGF$BioGasData

# namedBGF
namedBGF$BioGasData

```

Both contain only `NA` values.
This is because during object creation with `BGF()` no data exists to be filled in the `BioGasData`-layer, but as for each fermentation at least one observation in `BioGasData`-layer should exist, these empty lines are created.
They will be removed as soon as a `BGF` is updated with `update_BGF()`, which automatically happens during object creation with `from_standard_record()` or `from_AMPTSV2_report()`, or when a `BGF` is manipulated with dedicated functions of the `BGFanalyzeR` package.

Normally `update_BGF()` is not called by the package user directly, but here for illustration purpose:

```{r updated-BioGasData}
# remove empty columns in 'BioGasData'-layer
updatedBGF <- update_BGF(myBGF)

# inspect updated BGF
updatedBGF

# and its 'BioGasData'-layer
updatedBGF$BioGasData
```

The resulting `BGF` has no observations in the `BioGasData`-layer.

## Visualizing fermentation data

Building `BGF`'s with `BGF()` is nice but it objects created that way lack any experimental data. 
For the topic of how to get data into a `BGF`, several workflows exist, which all have a dedicated vignette ().

For the rest of this vignette, the focus is on how data visualization works with the `BGFanalyzeR` package.

To this end we need a `BGF` with data.
The package provides an example data set `LabscaleBiogas`, which is sufficient for this purpose.

Let's first inspect it:
```{r expBGF}
# bind the example data set to an object
expBGF <- LabscaleBiogas

# rename it
expBGF <- alter_whatever(expBGF,"ExpParam","name","Example BGF")

# inspect it
expBGF
```

This `BGF` is named '`r expBGF$ExpParam$name`' has a 15 fermentations.
Its `ExpParam`- and `metaData`-layer have 11 experimental parameters and meta variables, respectively.
In addition, it has 735 observations of 7 fermentation variables in its `BioGasData`-layer.

The rest of the listing is a bit different from the listings of 'myBGF' and 'namedBGF'.
Instead of the line `` `A '$yield' is not calculated yet` `` a tabular so called 'yield summary' is shown.

This summary is stored in the `metaData`-layer of a `BGF` and usually generated using `summarize_yield()`.
It contains information on the (final) `yield`, max `prduction` and the moment when max `production` occurred per user-specified reactor layout. If one layout was specified for more than one fermentation, a standard deviation for these values is also calculated.

### The product curve
But now, let's finally start visualizing out `BGF`.
The easiest way to do this, is simply `plot()` it.

```{r plot-BGF,fig.width=7,fig.height=5}
# plot the 'BGF'
plot(expBGF)
```

In this plot, the values of `BioGasData$product` are shown on the y-axis and `BioGasData$time` is on the x-axis. 
Fermentations are colored by their identifier in `BioGasData$reactor`.

This is the minimal information needed to compare biogas fermentations and thus this plot can be created as soon as data was entered to the `BioGasData`-layer .
The title contains `ExpParam$name`. A suitable x-axis label can be generated via the `xlab` argument in `plot()`. 

The unit 'Nml' in the y-axis label is read 'norm ml' and is a gas volume at 0°C and 1 atm pressure. This unit eventually needs correction if `BioGasData$product` has a different unit.

More advanced data visualization is possible using `bgf_plot()`.
With this function the data in `BioGasData$product`, `BioGasData$netProduct`, `BioGasData$production`, `BioGasData$rel_production` and `BioGasData$yield` can be easily plotted:

```{r product-plot,fig.width=7,fig.height=5}
# product plot
bgf_plot(expBGF,type = "product")
```

This produces a similar output like `` `plot(expBGF)` ``.
However, values of `BioGasData$product` are visualized as color coded lines, not dots. 
In addition a unit in the x-axis label is automatically, if `ExpParam$timeScale` was added to the `BGF`.

### The net product per layout curve
By specifying 'product' in the `type` argument of `bgf_plot()` a similar plot as with `plot()` was produced.

Now we will see what happens if we set `type` to 'netProduct':

```{r net-product-plot,fig.width=7,fig.height=5}
# net product by layout plot
bgf_plot(expBGF,type = "netProduct")
```

A new plot is generated!
Like in the previous plot, `BioGasData$time` is on the x-axis, but the y-axis scales with `BioGasData$net_product` and only eight instead of fifteen colored lines are plotted.

This is because each line in this plot belongs to a distinct reactor layout, not a fermentation!
Consequently, the values on the y-axis are mean values of `BioGasData$net_product` calculated over all fermentations that share the same layout.

We can check the reactor layout of a `BGF` using `get_ReactorLayout()`:

```{r get_reactorLayout}
# print reactor layout of expBGF
get_ReactorLayout(expBGF,feedback = TRUE)

# check how many fermentations
length(get_ReactorLayout(expBGF))

# check how many distinct layouts exist in the BGF
length(levels(get_ReactorLayout(expBGF)))
```

We can see, the `BGF`'s name, '`r expBGF$ExpParam$name`', and it's 'ReactorLayout', which has eight levels. 
Each level except for 'Cellulose' belongs to two fermentations, so the total length of the layout is fifteen.

Now that we understood what the differences between the `type` 'product' and 'netProduct' plots produced by `bgf_plot()` is, let's move on to the next type.

### The production curve
When using 'production' in the `type` argument of `bgf_plot()`, a curve is generated using `BioGasData$production`.


```{r production-plot,fig.width=7,fig.height=5}
# production plot
bgf_plot(expBGF,type = "production")
```

For each fermentation, the values of `BioGasData$prodution` are plotted as a colored line.
The y-axis is labeled as 'raw exhaust gas flow', because in a biogas fermentation the 'production' usually resembles a volume produced over time.
In case of our example it would have the unit '[Nml/d]' (read 'norm milliliter per day').
Thus, this plot displays how much biogas was produced per day for each fermentation.

### The relative production curve
The next `type` presented is 'relProduction', which produces also a line graphic but with adjusted `BioGasData$rel_production` values on the y-axis.

```{r rel-production-plot,fig.width=7,fig.height=5}
# relative product plot
bgf_plot(expBGF,type = "relProduction")
```

In `BioGasData$rel_production`, (fermentation-specific) the share of `BioGasData$product` at each time of the max `BioGasData$product` in '\%' is calculated.
Subtracting this value from 100\% results in the curves plotted above, from which the remaining gas production (in '\%') of each fermentation at each time can be read.

### yield plots

The last two plots presented in this vignette are a bit different from the previously shown, as they are made to visualize the 'yield summary' that is shown when a `BGF` is `print()`ed to the R console.

For the first yield plot, one can set `type` to 'yield_col'. 
This plot is directly build from the `metaData`-layer, like the 'yield summary' of the `print()` result and thus such a summary has to be created using `summarize_yield()` for this plot type to be working!

```{r yield-col-plot,fig.width=7,fig.height=5}
# yield col plot
bgf_plot(expBGF,type = "yield_col")
```

If such a 'yield summary' is present in `metaData`, a column plot is generated.
It has reactor layouts on the x-axis and a biogas yield on the y-axis.
The biogas yield of a fermentation is the amount of biogas produced from a defined amount of organic input material.
In our example the unit would be 'Nml/g~oTS~' (read 'norm milliliter per gram organic total solutes').

The height of each column resembles the mean yield calculated for a given layout. 
Error bars on top of each column illustrate the standard deviation around this mean value.
Error bars are only generated if at least two fermentations share the same reactor layout.


Lastly, if 'yield_box' is specified in the `type` argument of `bgf_plot`, a boxplot of `BioGasData$yield` values is generated for each reactor layout.

```{r yield-box-plot,fig.width=7,fig.height=5}
# yield box plot
bgf_plot(expBGF,type = "yield_box")
```

This plot is advantageous to the column plot, as it offers more flexibility in value selection, shows more details on the data and can be created without generating a 'yield summary' in `metaData`.

It is possible to omit the `type` argument when calling `bgf_plot()`. 
In this case the default value 'all' is used and all of the above mentioned plots are created at once.

## Further information
As mentioned above, different workflows of getting data into a `BGF` exist, which all have their own vignettes.
For educational purpose it is recommended to read at least the vignettes for AMPTS II-generated, and standard report-based `BGF`'s as this will build a holistic impression of the `BGFanalyzeR` package.

Furthermore, the vignette on manual `BGF`'s provides a deeper impression and shows how the package can help planing and setting up real world experiments.

In addition to the vignettes, the `BGFanalyzeR` package also provides a detailed documentation on each function.
If still open questions remain, don't hesitate in contacting the authors.
